A Bioconductor workflow for processing and analysing spatial proteomics data

### 2018-04-11 bioc-workflow-resubmit.Rnw
verison for resubmission

While we focus here on a LOPIT-type dataset, these analyses are
\hl{relevent} for any quantitative spatial proteomics data,  
irrespective of the fractionation (i.e. density gradient or 
differential centrifugation \cite{Itzhak:2016}) or quantitation 
(i.e. labelled or label-free) methods.

@@

There are two aspects related to data normalisation that are \hl{relevent}
to spatial proteomics data processing. The first one focuses on
reducing purely technical variation between channels without affecting
biological variability (i.e. the shape of the quantitatives
profiles). This normalisation will depend on the underlying
quantitative technology and the experimental design, and will not be
addressed in this workflow. The second aspect, and more specific to
spatial proteomics data, is scaling all the organelle-specific
profiles into \hl{the} same intensity interval (typically 0 and 1) by, for
example, dividing each intensity by the sum of the intensities for
that quantitative feature. This is not necessary in this example as
the intensities for each replicate have already been re-scaled to 1 in
Proteome Discoverer v1.4 Thermo Fisher. However, if the data require
normalisation, the user can execute the \texttt{normalise} function as
demonstrated in the below code chunk.

@@

Missing data are a recurrent issue in mass spectrometry applications,
and should be addressed independently of this workflow
\cite{Webb-Robertson:2015,Lazar:2016}. In \cite{Gatto:2014b}, we have
described how a high content in missing values in spatial proteomics
data and their inappropriate handling leads to a reduction of
sub-cellular resolution. \st{Missing data can be imputated} \hl{We can impute missing data} using
\Biocpkg{MSnbase}'s \texttt{impute} function. The method underlying
the imputation method is then determined by a \texttt{methods}
parameter (see \texttt{?impute} for available options). To impute
missing values using nearest neighbour imputation, one would

@@

\begin{figure}[!ht]
  \centering
<<namap, out.width=".65\\textwidth">>=
image2(is.na(combined), col = c("black", "white"),
       main = "Missing values (white cells) after combining replicates")
@  
\caption{Heatmap of missing values. Note that the features are
  re-ordered to highlight \hl{clusters} of proteins with similar numbers of
  missing values.}
  \label{fig:namap}
\end{figure}

@@

Having added the mouse markers to our \texttt{fData} from the
\texttt{pRolocmarkers}, we can now visualise these annotations on
the PCA plot using the \texttt{plot2D} function and then use the
\texttt{addLegend} function to map the marker classes to the
pre-defined colours. \hl{As previously mentioned, PCA transforms the
original high dimensional data into a set of linearly uncorrelated
principal components (PCs) such that the first accounts for as much
variability in the data as possible and each succeeding component in
turn has the highest variance possible under the constraint that it
be orthogonal to the preceding components. We saw in the previous
section how visualisation of the PCs is useful for quality control and checking
organelle seperation. Adding marker definiton allows one to quickly
see if known residents appear in defined clusters. One must be
careful though as different organelles may be resolved in different
dimensions.} For example, we can display the data along the first and
seventh PCs using the \texttt{dims} argument. Note that in these
calls to the \texttt{plot2D} function, we have omitted the
\texttt{fcol} argument and so by default the feature variable named
\texttt{"markers"} is used to annotate the plot. We choose to display PCs 1
and 7 to illustrate that while upper principal components explain
much less variability in the data (2.23\% for PC7, as opposed to
48.41\% for PC1), we see that the mitochondrial (purple) and
peroxisome (dark blue) clusters can be differentiated, despite the
apparent overlap in the visualisation of the two first PCs (Figure
\ref{fig:plotmarkers}).

@@

\hl{This is further highlighted if we plot the
profiles of these two clusters using the \texttt{plotDist} function}
(Figure \ref{fig:plotDist2}). \hl{The \texttt{plotDist} function is
another useful visualisation that relies on marker annotation. It
allows one to represent the protein profiles occupancy along the
gradient. While the PCA
plot enables efficient visualisation of the complete dataset and
assessment the relative separation of different sub-cellular niches,
comparing profiles of a few marker clusters is useful to assess how
exactly they differ (in terms of peak channels, for example). On
figure} \ref{fig:plotDist2}, \hl{we plot the profiles of the 
mitochondrial and peroxisome markers to highlight the differences 
in the channels labelled with tag 129C, also represented above  
along the 7th PC on the PCA plot on figure} \ref{fig:plotmarkers}.

@@

\subsection*{Features of interest}
In addition to adding annotation using the \texttt{addMarkers}
function, one can store specific sets of proteins by using the
\textit{Features of interest} infrastructure from the
\Biocpkg{MSnbase} package. If users have specific subsets of proteins
they wish to highlight in their data (possibly across multiple
experiments) they would first create a \texttt{FeaturesOfInterest}
object \hl{and then use the \textit{highlightOnPlot} function to 
visualise these}. For example, if we wanted to highlight 
proteins with the accession numbers Q8CG48, Q8CG47, Q8K2Z4, and 
Q8C156, which are some of the proteins known to form part of the 
13S condensin complex, we \hl{would call the following code:}

@@

The more \hl{relevent} comparison unit is not a single channel, but rather
the complete protein occupancy profiles, which are best visualised 
experiment-wide on a PCA plot. As such, we prefer to focus on the
direct, qualitative comparison of individual replicate PCA plots,
assuring that each displays acceptable sub-cellular resolution. Note
that in the code chunk below, we mirror the x-axis to represent the
two figures with the same orientation. The interactive "compare"
application part of the \Biocpkg{pRolocGUI} package is also useful for
examining replicate experiments (see the next section interactive
visualisation for details).

@@

\hl{The above analysis is computationally intensive and best parallelised over
multiple workers. This phenoDisco analysis took ~24 hours to complete
when parallelised over 40 workers. As such, in the interest of time 
users can access the above results which are
pre-computed and stored along with the \texttt{pRolocdata} package. 
Please see the appendix to load these results.}

@@

There are several classification algorithms readily available in
\Biocpkg{pRoloc}, which are documented in the dedicated
\href{http://bioconductor.org/packages/release/bioc/vignettes/pRoloc/inst/doc/pRoloc-ml.pdf}{\Biocpkg{pRoloc}
  machine learning techniques vignette}.  We find the general \hl{tendency}
to be that it is not the choice of classifier, but the improper
optimisation of the algorithmic parameters, that limits classification
accuracy. Before employing any classification algorithm and generating
a model on the training data, one must first find the optimal
parameters for the algorithm of choice.

@@

\hl{In the code chunk below we then pass these weights 
to the \texttt{svmOptimisation} function. Once again, we provide
the optimisation results for users to load directly if they wish to save
computational time, please see the appendix to load these results.}

@@

There are many ways to set thresholds and the choice of method will
depend on the biological question and experimental design at hand. One
viable approach in the frame of the above experimetal design would be
to manually set a FDR, say 5\%, per organelle. To do this the user
would examine the top scoring predictions for each organelle, and then
set a threshold at the score at which they achieve 5\% of false
assignments per organelle.  The definintion of a false assignment
would depend on the information available, for example, validity or
lack of validity for the localisation from another experiment as
reported in the literature or a reliable database.  If such
information is not available, one crude method is to set a threshold
per organelle by extracting the median or 3rd \hl{quartile} score per
organelle.  For example, in the code chunk below, we use the
\texttt{orgQuants} function to extract the median organelle scores and
then pass these scores to the \texttt{getPredictions} function to
extract the new localisations that meet this scoring criteria. Any
sub-cellular predictions that fall below the specified thresholds are
labelled as unknown.

@@

Now we have contructed the query parameters we can use the
\texttt{makeGoSet} function to retrieve and build an auxiliary GO
\texttt{MSnSet} as described above. By default, the cellular component
terms are downloaded, without any filtering on evidence codes. It is
also possible to download terms from the molecular function and
biological process GO namespaces, and also apply filtering based on
evidence codes as desired, see \texttt{?makeGoSet} for more details. 
\hl{(We also provide the pre-computed \texttt{gocc} object for users
if they wish to load directly, please see the appendix).}

@@

The first step, as with any machine learning algorithm, is to optimise
any free paramaters of the classifier. For the $k$-NN TL classifier
there are two sets of parameters that need optimising: the first set
are the $k$'s for the primary and auxiliary data sources required for
the nearest neighbour calculations for each data source. The second
set of parameters (noted by a vector of $\theta$ weights) that require
optimising are the class weights, one per subcellular niche, that
control the proportion of primary and auxiliary data to use for
learning. A weight can take any real value number between 0 and 1. A
weight of $\theta = 1$ indicates that all weight is given to the
primary data (and this implicitly implies that a weight of
$1 - \theta = 0$ is given to the auxiliary data), and similarly a
weight of $\theta = 0$ implies that all weight is given to the
auxiliary data (so 0 is given to the primary source). If we conduct a
parameter \hl{search} and test weights $\theta = {0, 1/3, 2/3, 1}$ for each
class, and if we have, for example 10 subcellular niches, this will
result in \texttt{4$^{10}$} different combinations of parameters to
test. The parameter optimisation is therefore time consuming and as
such we recommend making use of a computing cluster (code and
submissing scripts are also available in the supporting
information). The markers in the \texttt{hl} dataset contain 14
subcellular classes. If we examine these markers and classes on the
PCA plot above we can see that in particular the two ribosomes and two
nuclear compartments are highly separated along the first two
components, this is also evident from the profiles plot which gives us
a good indication that these subcellular niches are well-resolved in
the hyperLOPIT dataset. Transfer learning is particularly useful for
classes that are not as well separated. We find that subcellular
niches that are well-separated under hyperLOPIT and LOPIT obtain a
class score of 1 (i.e. use only primary data from transfer learning
\cite{Breckels:2016}). Therefore, for the optimisation stage of the
analyses we can already infer a subcellular class weight of 1 for
these niches and only optimise over the remaining organelles.  This
can significantly cut down optimisation time as by removing these 4
classes from the optimisation (and not the classification) we only
have \texttt{4$^{10}$} class weight combinations to consider instead
of \texttt{4$^{14}$} combinations.

In the example below we \hl{first} remove these 4 classes from the marker set
\hl{and create a new marker set called tlmarkers. Then we}
re-run the \texttt{knnOptimisation} for each data source and then run
the \texttt{knntlOptimisation} with the 10 remaining classes. (Note:
this is not run live as the \texttt{hl} dataset with 10 classes,
707 markers and 4$^{10}$ combinations of parameters takes around 76
hours to run on the University of Cambridge HPC using 256 workers. 
\hl{To save time for users, the results of the following optimisation 
are pre-computed and provided with the \texttt{pRolocdata} package. 
Please see the appendix for details on how
to load these directly}.)q

@@

From examining the parameter \hl{search} plots as described in section
\textit{Optimisation}, we find the best $k$'s for both the primary and
auxiliary are 3.

@@

We generally find supervised learning more suited to the task of
protein localisation prediction in which we use high-quality curated
marker proteins to build a classifier, instead of using an entirely
unsupervised approach to look for clusters and then look for
enrichment of organelles and complexes. In the latter we do not make
good use of valuable prior knowledge, and in our experience
unsupervised clustering can be extremely difficult due to \hl{
(i) the loose definition of what constitutes a cluster (for example 
whether it is defined by the quantitative data or the localisation 
information), (ii) the influence of the algorithm assumption on the 
cluster identification (for example parametric or non-parametric) 
and (iii) poor estimates of the number of clusters that may appear 
in the data.}

@@

Exporting to a spreadsheet however loses a lot of important
information, such as the processing data, and the sample metadata in
the \textit{phenoData} slot. Other objects, such as parameters from the
machine learning optimisation, cannot be represented as tabular
data. To directly serialise R objects to disk, on can use the standard
\texttt{save} function, and later reload the object using \st{save}
\texttt{load}. For example, to save and then re-load the parameters
from the SVM optimisation,

>>>>

While we focus here on a LOPIT-type dataset, these analyses are
relevent for any quantitative spatial proteomics data,  
irrespective of the fractionation (i.e. density gradient or 
differential centrifugation \cite{Itzhak:2016}) or quantitation 
(i.e. labelled or label-free) methods.

@@

There are two aspects related to data normalisation that are relevent
to spatial proteomics data processing. The first one focuses on
reducing purely technical variation between channels without affecting
biological variability (i.e. the shape of the quantitatives
profiles). This normalisation will depend on the underlying
quantitative technology and the experimental design, and will not be
addressed in this workflow. The second aspect, and more specific to
spatial proteomics data, is scaling all the organelle-specific
profiles into the same intensity interval (typically 0 and 1) by, for
example, dividing each intensity by the sum of the intensities for
that quantitative feature. This is not necessary in this example as
the intensities for each replicate have already been re-scaled to 1 in
Proteome Discoverer v1.4 Thermo Fisher. However, if the data require
normalisation, the user can execute the \texttt{normalise} function as
demonstrated in the below code chunk.

@@

Missing data are a recurrent issue in mass spectrometry applications,
and should be addressed independently of this workflow
\cite{Webb-Robertson:2015,Lazar:2016}. In \cite{Gatto:2014b}, we have
described how a high content in missing values in spatial proteomics
data and their inappropriate handling leads to a reduction of
sub-cellular resolution. We can impute missing data using
\Biocpkg{MSnbase}'s \texttt{impute} function. The method underlying
the imputation method is then determined by a \texttt{methods}
parameter (see \texttt{?impute} for available options). To impute
missing values using nearest neighbour imputation, one would

@@

\begin{figure}[!ht]
  \centering
<<namap, out.width=".65\\textwidth">>=
image2(is.na(combined), col = c("black", "white"),
       main = "Missing values (white cells) after combining replicates")
@  
\caption{Heatmap of missing values. Note that the features are
  re-ordered to highlight clusters of proteins with similar numbers of
  missing values.}
  \label{fig:namap}
\end{figure}

@@

Having added the mouse markers to our \texttt{fData} from the
\texttt{pRolocmarkers}, we can now visualise these annotations on
the PCA plot using the \texttt{plot2D} function and then use the
\texttt{addLegend} function to map the marker classes to the
pre-defined colours. As previously mentioned, PCA transforms the
original high dimensional data into a set of linearly uncorrelated
principal components (PCs) such that the first accounts for as much
variability in the data as possible and each succeeding component in
turn has the highest variance possible under the constraint that it
be orthogonal to the preceding components. We saw in the previous
section how visualisation of the PCs is useful for quality control and checking
organelle seperation. Adding marker definiton allows one to quickly
see if known residents appear in defined clusters. One must be
careful though as different organelles may be resolved in different
dimensions. For example, we can display the data along the first and
seventh PCs using the \texttt{dims} argument. Note that in these
calls to the \texttt{plot2D} function, we have omitted the
\texttt{fcol} argument and so by default the feature variable named
\texttt{"markers"} is used to annotate the plot. We choose to display PCs 1
and 7 to illustrate that while upper principal components explain
much less variability in the data (2.23\% for PC7, as opposed to
48.41\% for PC1), we see that the mitochondrial (purple) and
peroxisome (dark blue) clusters can be differentiated, despite the
apparent overlap in the visualisation of the two first PCs (Figure
\ref{fig:plotmarkers}).

@@

This is further highlighted if we plot the
profiles of these two clusters using the \texttt{plotDist} function
(Figure \ref{fig:plotDist2}). The \texttt{plotDist} function is
another useful visualisation that relies on marker annotation. It
allows one to represent the protein profiles occupancy along the
gradient. While the PCA
plot enables efficient visualisation of the complete dataset and
assessment the relative separation of different sub-cellular niches,
comparing profiles of a few marker clusters is useful to assess how
exactly they differ (in terms of peak channels, for example). On
figure \ref{fig:plotDist2}, we plot the profiles of the 
mitochondrial and peroxisome markers to highlight the differences 
in the channels labelled with tag 129C, also represented above  
along the 7th PC on the PCA plot on figure \ref{fig:plotmarkers}.

@@

\subsection*{Features of interest}
In addition to adding annotation using the \texttt{addMarkers}
function, one can store specific sets of proteins by using the
\textit{Features of interest} infrastructure from the
\Biocpkg{MSnbase} package. If users have specific subsets of proteins
they wish to highlight in their data (possibly across multiple
experiments) they would first create a \texttt{FeaturesOfInterest}
object and then use the \textit{highlightOnPlot} function to 
visualise these. For example, if we wanted to highlight 
proteins with the accession numbers Q8CG48, Q8CG47, Q8K2Z4, and 
Q8C156, which are some of the proteins known to form part of the 
13S condensin complex, we would call the following code:

@@

The more relevent comparison unit is not a single channel, but rather
the complete protein occupancy profiles, which are best visualised 
experiment-wide on a PCA plot. As such, we prefer to focus on the
direct, qualitative comparison of individual replicate PCA plots,
assuring that each displays acceptable sub-cellular resolution. Note
that in the code chunk below, we mirror the x-axis to represent the
two figures with the same orientation. The interactive "compare"
application part of the \Biocpkg{pRolocGUI} package is also useful for
examining replicate experiments (see the next section interactive
visualisation for details).

@@

The above analysis is computationally intensive and best parallelised over
multiple workers. This phenoDisco analysis took ~24 hours to complete
when parallelised over 40 workers. As such, in the interest of time 
users can access the above results which are
pre-computed and stored along with the \texttt{pRolocdata} package. 
Please see the appendix to load these results.

@@

There are several classification algorithms readily available in
\Biocpkg{pRoloc}, which are documented in the dedicated
\href{http://bioconductor.org/packages/release/bioc/vignettes/pRoloc/inst/doc/pRoloc-ml.pdf}{\Biocpkg{pRoloc}
  machine learning techniques vignette}.  We find the general tendency
to be that it is not the choice of classifier, but the improper
optimisation of the algorithmic parameters, that limits classification
accuracy. Before employing any classification algorithm and generating
a model on the training data, one must first find the optimal
parameters for the algorithm of choice.

@@

In the code chunk below we then pass these weights 
to the \texttt{svmOptimisation} function. Once again, we provide
the optimisation results for users to load directly if they wish to save
computational time, please see the appendix to load these results.

@@

There are many ways to set thresholds and the choice of method will
depend on the biological question and experimental design at hand. One
viable approach in the frame of the above experimetal design would be
to manually set a FDR, say 5\%, per organelle. To do this the user
would examine the top scoring predictions for each organelle, and then
set a threshold at the score at which they achieve 5\% of false
assignments per organelle.  The definintion of a false assignment
would depend on the information available, for example, validity or
lack of validity for the localisation from another experiment as
reported in the literature or a reliable database.  If such
information is not available, one crude method is to set a threshold
per organelle by extracting the median or 3rd quartile score per
organelle.  For example, in the code chunk below, we use the
\texttt{orgQuants} function to extract the median organelle scores and
then pass these scores to the \texttt{getPredictions} function to
extract the new localisations that meet this scoring criteria. Any
sub-cellular predictions that fall below the specified thresholds are
labelled as unknown.

@@

Now we have contructed the query parameters we can use the
\texttt{makeGoSet} function to retrieve and build an auxiliary GO
\texttt{MSnSet} as described above. By default, the cellular component
terms are downloaded, without any filtering on evidence codes. It is
also possible to download terms from the molecular function and
biological process GO namespaces, and also apply filtering based on
evidence codes as desired, see \texttt{?makeGoSet} for more details. 
(We also provide the pre-computed \texttt{gocc} object for users
if they wish to load directly, please see the appendix).

@@

The first step, as with any machine learning algorithm, is to optimise
any free paramaters of the classifier. For the $k$-NN TL classifier
there are two sets of parameters that need optimising: the first set
are the $k$'s for the primary and auxiliary data sources required for
the nearest neighbour calculations for each data source. The second
set of parameters (noted by a vector of $\theta$ weights) that require
optimising are the class weights, one per subcellular niche, that
control the proportion of primary and auxiliary data to use for
learning. A weight can take any real value number between 0 and 1. A
weight of $\theta = 1$ indicates that all weight is given to the
primary data (and this implicitly implies that a weight of
$1 - \theta = 0$ is given to the auxiliary data), and similarly a
weight of $\theta = 0$ implies that all weight is given to the
auxiliary data (so 0 is given to the primary source). If we conduct a
parameter search and test weights $\theta = {0, 1/3, 2/3, 1}$ for each
class, and if we have, for example 10 subcellular niches, this will
result in \texttt{4$^{10}$} different combinations of parameters to
test. The parameter optimisation is therefore time consuming and as
such we recommend making use of a computing cluster (code and
submissing scripts are also available in the supporting
information). The markers in the \texttt{hl} dataset contain 14
subcellular classes. If we examine these markers and classes on the
PCA plot above we can see that in particular the two ribosomes and two
nuclear compartments are highly separated along the first two
components, this is also evident from the profiles plot which gives us
a good indication that these subcellular niches are well-resolved in
the hyperLOPIT dataset. Transfer learning is particularly useful for
classes that are not as well separated. We find that subcellular
niches that are well-separated under hyperLOPIT and LOPIT obtain a
class score of 1 (i.e. use only primary data from transfer learning
\cite{Breckels:2016}). Therefore, for the optimisation stage of the
analyses we can already infer a subcellular class weight of 1 for
these niches and only optimise over the remaining organelles.  This
can significantly cut down optimisation time as by removing these 4
classes from the optimisation (and not the classification) we only
have \texttt{4$^{10}$} class weight combinations to consider instead
of \texttt{4$^{14}$} combinations.

In the example below we first remove these 4 classes from the marker set
and create a new marker set called tlmarkers. Then we
re-run the \texttt{knnOptimisation} for each data source and then run
the \texttt{knntlOptimisation} with the 10 remaining classes. (Note:
this is not run live as the \texttt{hl} dataset with 10 classes,
707 markers and 4$^{10}$ combinations of parameters takes around 76
hours to run on the University of Cambridge HPC using 256 workers. 
To save time for users, the results of the following optimisation 
are pre-computed and provided with the \texttt{pRolocdata} package. 
Please see the appendix for details on how
to load these directly.

@@

From examining the parameter search plots as described in section
\textit{Optimisation}, we find the best $k$'s for both the primary and
auxiliary are 3.

@@

We generally find supervised learning more suited to the task of
protein localisation prediction in which we use high-quality curated
marker proteins to build a classifier, instead of using an entirely
unsupervised approach to look for clusters and then look for
enrichment of organelles and complexes. In the latter we do not make
good use of valuable prior knowledge, and in our experience
unsupervised clustering can be extremely difficult due to
(i) the loose definition of what constitutes a cluster (for example 
whether it is defined by the quantitative data or the localisation 
information), (ii) the influence of the algorithm assumption on the 
cluster identification (for example parametric or non-parametric) 
and (iii) poor estimates of the number of clusters that may appear 
in the data.

@@

Exporting to a spreadsheet however loses a lot of important
information, such as the processing data, and the sample metadata in
the \textit{phenoData} slot. Other objects, such as parameters from the
machine learning optimisation, cannot be represented as tabular
data. To directly serialise R objects to disk, on can use the standard
\texttt{save} function, and later reload the object using
\texttt{load}. For example, to save and then re-load the parameters
from the SVM optimisation,
